---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
author: "Sriharsha M S"
html_document: "PA1_template.html"
---

It is now possible to collect a large amount of data about personal
movement using activity monitoring devices such as a
[Fitbit](http://www.fitbit.com), [Nike
Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or
[Jawbone Up](https://jawbone.com/up). These type of devices are part of
the "quantified self" movement -- a group of enthusiasts who take
measurements about themselves regularly to improve their health, to
find patterns in their behavior, or because they are tech geeks. But
these data remain under-utilized both because the raw data are hard to
obtain and there is a lack of statistical methods and software for
processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded here: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there
are a total of 17,568 observations in this
dataset.


### Basic settings
Basic settings for Project.
```{r}
echo = TRUE  # Always make code visible
options(scipen = 999)  # Turn off scientific notations for numbers
```

## Loading and preprocessing the data
Unzip and load Activity monitoring data 
```{r}
unzip("activity.zip")
activityData <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
activityData$month <- as.numeric(format(activityData$date, "%m"))
activityData_noNA <- na.omit(activityData)
rownames(activityData_noNA) <- 1:nrow(activityData_noNA)
head(activityData_noNA)
dim(activityData_noNA)
library(ggplot2)
```


## What is mean total number of steps taken per day?
For this part of the assignment, we can ignore the missing values in the dataset.

* Make a histogram of the total number of steps taken each day
```{r}
ggplot(activityData_noNA, aes(date, steps)) + geom_bar(stat = "identity", colour = "red", fill = "red", width = 0.5) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")
```

* Calculate and report the mean and median total number of steps taken per day

Mean total number of steps taken per day:
```{r}
totalSteps <- aggregate(activityData_noNA$steps, list(Date = activityData_noNA$date), FUN = "sum")$x
mean(totalSteps)
```
Median total number of steps taken per day:
```{r}
median(totalSteps)
```


## What is the average daily activity pattern?
* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
avgSteps <- aggregate(activityData_noNA$steps, list(interval = as.numeric(as.character(activityData_noNA$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"

ggplot(avgSteps, aes(interval, meanOfSteps)) + geom_line(color = "red", size = 0.6) + labs(title = "Time Series Plot of the 5-minute Interval", x = "5-minute intervals", y = "Average Number of Steps Taken")
```

* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
avgSteps[avgSteps$meanOfSteps == max(avgSteps$meanOfSteps), ]
```

## Imputing missing values
* The total number of rows with NAs:

```{r}
sum(is.na(activityData))
```

* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, we could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Our strategy is to use the mean for that 5-minute interval to fill each NA value in the steps column.

* Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
newActivityData <- activityData 
for (i in 1:nrow(newActivityData)) {
    if (is.na(newActivityData$steps[i])) {
        newActivityData$steps[i] <- avgSteps[which(newActivityData$interval[i] == avgSteps$interval), ]$meanOfSteps
    }
}

head(newActivityData)
sum(is.na(newActivityData))
```

* Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
ggplot(newActivityData, aes(date, steps)) + geom_bar(stat = "identity",
                                             colour = "red",
                                             fill = "red",
                                             width = 0.5) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day (no missing data)", x = "Date", y = "Total number of steps")
```

* Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Mean total number of steps taken per day:
```{r}
newTotalSteps <- aggregate(newActivityData$steps, 
                           list(Date = newActivityData$date), 
                           FUN = "sum")$x
newMean <- mean(newTotalSteps)
newMean
```
Median total number of steps taken per day:
```{r}
newMedian <- median(newTotalSteps)
newMedian
```
Compare them with the two before imputing missing data:
```{r}
oldMean <- mean(totalSteps)
oldMedian <- median(totalSteps)
newMean - oldMean
newMedian - oldMedian
```
So, after imputing the missing data, the new mean of total steps taken per day is the same as that of the old mean; the new median of total steps taken per day is greater than that of the old median.


## Are there differences in activity patterns between weekdays and weekends?
* Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
head(newActivityData)
newActivityData$weekdays <- factor(format(newActivityData$date, "%A"))
levels(newActivityData$weekdays)
levels(newActivityData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(newActivityData$weekdays)
table(newActivityData$weekdays)
```

* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
avgSteps <- aggregate(newActivityData$steps, 
                      list(interval = as.numeric(as.character(newActivityData$interval)), 
                           weekdays = newActivityData$weekdays),
                      FUN = "mean")
names(avgSteps)[3] <- "meanOfSteps"
library(lattice)
xyplot(avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```

As demonstrated above, activity patterns vary between weekdays and weekend days. This suggests that the imputation strategy could be refined by using an average value for the interval for the subset of data for weekdays or weekend days accordingly.
